Analysis-ready VCF at Biobank scale using Zarr

### 2025-01-31 response-to-reviewers.tex

The authors' benchmarks overall show that VCF Zarr has superior compression and
computational analysis performance at scale relative to data stored as
row-oriented VCF and that VCF Zarr is competitive with specialized storage
solutions that require similarly specialized tools and access libraries for
querying. An attractive feature is that VCF Zarr allows for variant annotation
workflows that do not require full dataset copy and conversion. Another key
point is that Zarr is a high-level spec and data model for the chunked storage
of n-d arrays, rather than a byte-level encoding designed specifically around
the genomic variant data type. I personally have used Zarr productively for
several applications unrelated to statistical genetics. While Zarr VCF mildly
underperforms some of the specialized formats (Savvy in compute, Genozip in
compression) in a few instances, I believe the accessibility, interoperability,
and reusability gains of Zarr make the small tradeoff well worthwhile.

Because Zarr has seen heavy adoption in other scientific communities like the
geospatial and Earth sciences, and is well integrated in the scientific Python
stack, I think it holds potential for greater reusability across the ecosystem.
As such, I think the VCF Zarr spec is a highly valuable if not overdue
contribution to an entrenched field that has recently been confronted by a
scalability wall.

Overall, the paper is clear, comprehensive, and well written. 
\end{point}
\begin{reply}
Thank you!  We are delighted that you agree with us on the transformative
potential of VCF Zarr, and thank you for your insightful points and
helpful suggestions.
\end{reply}

\begin{point}
The benefits for large scientific datasets to be analysis-ready
cloud-optimized (ARCO) have been well articulated by Abernathey et al., 2021.
However, I do think that the "local"/HPC single-file use case is still
important and won't disappear any time soon, and for some file system use
cases, expansive and deep hierarchies can be performance limiting (this was
hinted at in one of the benchmarks). In this scenario would a large Zarr VCF
perform reasonably well (or even better on some file systems) via a single
local zip store?
\end{point}
\begin{reply}
This is a good point, and we have added several mentions of the Zip file backend
as well as benchmarks using it. We added a new paragraph on pg5, (l318-338)
to discuss this point. We also added benchmarks based on the SARS-CoV-2
dataset, which is distributed as a Zipfile.
\end{reply}

@@

\begin{point}
The authors discuss the possibility of storing an index mapping genomic
coordinates to chunk indexes. Have Zarr-based formats in other fields like
geospatial introduced their own indexing approaches to take inspiration from?
\end{point}
\begin{reply}
We have added a new \texttt{region\_index} field for this purpose, which 
supports efficient genomic range searches. This is discussed in the last 
paragraph of the vcf2zarr Methods section (pg 13).
\end{reply}

\begin{point}
Since VCF Zarr is still a draft proposal, it could be useful to indicate
where community discussions are happening and how potential new contributors
can get involved, if possible. This doesn't need to be in the paper per se, but
perhaps documented in the spec repo.
\end{point}
\begin{reply}
Thank you, that is an excellent idea. We have added basic guidance for 
contributors to the specification repo.
\end{reply}

\begin{point}
In the background: "For the representation to be FAIR, it must also be
accessible," -- A is for "accessible", so "also" doesn't make sense.
\end{point}
\begin{reply}
Fixed
\end{reply}

\begin{point}
"There is currently no efficient, FAIR representation...". Just a nit and
feel free to ignore, but the solution you present is technically "current".
\end{point}
\begin{reply}
You are right, but we would like to keep this as it is for narrative clarity.
\end{reply}

@@

\begin{point}
In Figure 2, the zarr line is occluded by the sav line and hard to see.
\end{point}
\begin{reply}
The benchmarks have changed slightly, and the lines are now more separated.
\end{reply}

\reviewersection
\section*{Reviewer 2 comments}

\begin{point}
The paper presents an encoding of the VCF data using Zarr to enable fast
retrieving subsets of the data. A vcf2arr conversion was provided and validated
on both simulated and real-world data sets. The topic of this work is
interesting and of good values, however, the experimental studies and
contributions should be considerable improved.
\end{point}
\begin{reply}
Thank you for your valuable feedback; it has resulted in a much improved
manuscript. We have significantly extended the scope 
of this work by providing 
\begin{itemize}
\item The vcztools program, which converts from Zarr to VCF efficiently.
\item A new case study on the Our Future Health genotype data, for 
651K samples.
\item A new case study on the All of Us exome-like data, for 245K
samples.
\item A new case study on Norway Spruce data, for 1063 samples
and 3.75 \emph{billion} variants.
\item A new case study on SARS-CoV-2 data for 4.3 million samples.
\item A new section demonstrating a prototype of the SAIGE software
incorporating support for VCF Zarr, with excellent performance.
\item A new section exploring the use of VCF Zarr in cloud computing 
platforms, showing how data processing rates of up to 25.5GiB/s
are possible.
\item A new section exploring the use of GPU acceleration on VCF Zarr data.
\end{itemize}
We hope that you agree that these extensive additions
and demonstrations on a selection of the largest genomic datasets
in the world (in terms of both sample and genome size),
demonstrate the real-world value and transformative potential of VCF Zarr.
\end{reply}

@@

\begin{point}
The proposed method is simply a conversion from VCF to Zarr format. Since
both are existing formats, the contributions and originality of this work are
not impressive.
\end{point}
\begin{reply}
We believe that the extensive work detailed in the previous response
addresses this criticism.
\end{reply}

\begin{point}
The compression and query performance is the main concern of this work. The
method should be compared with other state-of-the-art queriable VCF compressors
like GTC, GBC, and GSC. [References ommitted]
\end{point}
\begin{reply}
We disagree on this point, and respectfully point out that for our 
purposes, these methods are essentially equivalent to Genozip
(which we have included in our benchmarks as a representative 
of this class of method). 
As we argue in the ``Calculating with the genotype matrix'' section,
a fundamental point
is that compression methods that only provide access to data 
output via VCF text are limiting the rate at which the data can be 
processed to the rate at which VCF text can be parsed. Producing VCF text
as output cannot be seen as the end point of analysis, and the rate at which it can be
done (beyond a certain minimum value of around 120 MiB/s, as we argue in the 
new vcztools section) is of no real practical interest. Our benchmarks 
in e.g. Figures 3 and 4 show that the Genozip like approach 
of highly compressed genotype data that is accessed only via VCF text
cannot form the basis of efficient processing pipelines. It is 
intrinsically inefficient.

@@

We have included references to GTC, GBC and GSC, and discussing them in the 
``Storing genetic variation data'' section, directing the reader to 
the benchmarks performed  in the ``Calculating with the genotype matrix''
section.
\end{reply}

\begin{point}
The method should be evaluated on more real VCF data sets.
\end{point}
\begin{reply}
We have added four new case studies applying VCF Zarr
to some of the world's largest 
datasets with different data modalities (genotype, exome-like, whole genome sequence),
on three very different organisms (Human, Tree, Virus)
with widely differing genome properties. 
\end{reply}

### 2025-02-06 paper.tex
SC2 dataset detail updates

\textbf{Results:}
Zarr is a format for storing
multi-dimensional data that is widely used across the sciences,
and is ideally suited to massively parallel processing.
We present the VCF Zarr specification, an encoding of the
VCF data model using Zarr, along with fundamental software infrastructure
for efficient and reliable conversion at scale.
We show how this format is far more efficient than
standard VCF based approaches,
and competitive with specialised methods for
storing genotype data in terms of compression ratios
and single-threaded calculation performance.
We present case studies on subsets of three large 
human datasets (Genomics England: $n$=78,195;
Our Future Health: $n$=651,050; 
All of Us: $n$=245,394) along with whole genome
datasets for Norway Spruce ($n$=1,063) and 
SARS-CoV-2 ($n$=4,384,310).
We demonstrate the potential for VCF Zarr to enable a new generation
of high-performance and cost-effective applications
via illustrative examples using cloud computing and GPUs.

@@

In this article, we decouple the VCF data model from its row-oriented
file definition, and show how the data can be
compactly stored and efficiently analysed in a cloud-native, FAIR manner.
We do this by translating VCF data into Zarr format,
a method of storing large-scale multidimensional data as a regular
grid of compressed chunks.
Zarr's elegant simplicity and first-class support for
cloud object stores and parallel computing have led to
it gaining substantial traction
across the sciences, and it is now used in multiple petabyte-scale
datasets in cloud deployments (see Methods for details).
We present the VCF Zarr specification that formalises this
mapping, along with the \texttt{vcf2zarr} and \texttt{vcztools} 
utilities to reliably convert VCF to and from Zarr at scale.
We show that VCF Zarr is much more compact than
VCF and is competitive with state-of-the-art
file-based VCF compression tools.
Moreover, we show that Zarr's storage of data in an analysis-ready
format greatly facilitates computation,
with various benchmarks being substantially faster than
\texttt{bcftools} based pipelines, and again competitive
with state-of-the-art file-oriented methods. 
We demonstrate the scalability and flexibility of VCF Zarr
via case-studies on five boundary-pushing datasets.
We examine conversion and compression performance on three different
data modalities for large human datasets: whole genome sequence
data from Genomics England (5X smaller than gzipped VCF);
genotype data from Our Future Health (2.5X smaller than gzipped VCF);
and exome-like data from All of Us (1.1X larger than gzipped VCF). 
We then examine performance 
for species with large genomes using Norway Spruce data (3.2X smaller
than gzipped VCF),
and in a large collection of short whole genome alignments
using SARS-CoV-2 data 
(81X smaller than gzipped FASTA). 
Interoperability with the vast ecosystem of tools
based on VCF~\citep{danecek2021twelve,garrison2022spectrum}
is crucial. 
% TODO sentence too long
We first present the \texttt{vcztools} program,
which implements a subset of the functionality in \texttt{bcftools}
and allows data stored in Zarr at some URL to be 
treated as if it were a VCF or BCF file,
and demonstrate that it can be used as a practical replacement 
in processing pipelines.
We then present an illustrative example of how existing
methods can adopt VCF Zarr by modifying the popular 
SAIGE~\cite{zhou2018efficiently, zhou2020scalable} GWAS software
(written in C++), and show significant speedups over the existing 
VCF (7.9X) and BCF (3.4X) backends in a simple benchmark.
Finally, we illustrate the potential 
for a new generation of Zarr-based applications via some illustative
benchmarks. First, we demonstrate the very high levels of throughput 
performance available on public clouds when data in an object 
store is accessed in an asynchronous and highly parallel manner.
Finally, we demonstrate the ease and efficiency with which data
in Zarr format can be processed on a GPU, opening many exciting
avenues for future research.

@@

\subsection{Case study: SARS-CoV-2 alignment data}
The SARS-CoV-2 pandemic resulted in viral sequence data collection at
an unprecedented scale, with over 15 million whole genomes incorporated into
one resource as of 2024~\citep{hinrichs2024ongoing}. Data at such scale
overwhelmed classical phylogenetic methods, leading to the development
of a new class of tools~\citep{turakhia2021ultrafast,ye2022matoptimize,zhan2023towards}
capable of handling millions of samples.
The Viridian dataset~\citep{hunt2024addressing} is a consistently assembled
set of consensus sequences for 4,384,310 samples, addressing a range of systematic
errors. The \texttt{sc2ts} inference tool uses VCF-Zarr as its
input format, and Zhan et al have made the Viridian dataset,
aligned to the reference using MAFFT~\citep{katoh2013mafft},
available for download~\citep{zhan2023towards}. This dataset has quite
different properties to those examined in previous case studies, and
illustrates the flexibility of Zarr across a range of biological applications.

\begin{table}
\caption{Summary for VCF Zarr conversion of Viridian SARS-CoV-2
whole genome alignments and metadata 
(4,384,310 samples, 29903 variants)
consisting of 37 fields and 391 MiB of storage, 
of which 280MiB is used by alignments ($\sim 81$X smaller 
than gzipped FASTA). The chunk size is 100 variants $\times$
10,000 samples.
Shown are the top 9 fields consuming at least 1\% of the 
overall storage (see Table~\ref{tab-genomics-england-data} for column details).
Field names (derived from source TSV file) are shortened for 
display purposes by replacing ``Genbank'' with ``G''
and ``Viridian'' with ``V''.
\label{tab-sc2-data}}
\begin{tabular}{llS[table-format=3.2]S[table-format=3.2]S[table-format=3.2]}
\toprule
{Field} & {Type} & {Storage} & {Compress} & {\% Total} \\
\midrule
call\_genotype & int8 & 279.38 MiB & 447.5 & 80.04\% \\
sample\_G\_tree\_name & str & 13.18 MiB & 2.5 & 3.78\% \\
sample\_G\_accession & str & 7.31 MiB & 4.6 & 2.09\% \\
sample\_Sample & str & 7.26 MiB & 4.6 & 2.08\% \\
sample\_Experiment & str & 6.9 MiB & 4.8 & 1.98\% \\
sample\_id & str & 6.83 MiB & 4.9 & 1.96\% \\
sample\_V\_N & int16 & 4.32 MiB & 1.9 & 1.24\% \\
sample\_V\_pangolin\_1.29 & str & 3.98 MiB & 8.4 & 1.14\% \\
sample\_V\_pangolin & str & 3.89 MiB & 8.6 & 1.11\% \\
\bottomrule
\end{tabular}
\end{table}

@@

Decompressed, the Viridian dataset consists of 48
FASTA files totalling 125 GiB of storage and a 1.4 GiB
TSV file for sample metadata.
When aligned to the reference and converted to Zarr, the
entire dataset is stored in a single 391 MiB Zip file.
When compressed with bgzip (so they can be indexed; see next 
paragraph) the alignments require 22 GiB of storage.
Table~\ref{tab-sc2-data} summarises the storage used by individual
arrays, and shows that the alignment data requires 298 MiB,
around 81X smaller than gzipped FASTA.
The original dataset is distributed as xz compressed FASTA files, which
achieves even higher compression levels at a total of 218 MiB.
However, these files must be decompressed in order to access the data
using standard FASTA processing techniques.

Computational accessibility for the Zarr is excellent. For example,
computing the total number of missing or ambiguous bases for 
all 4.4 million samples required 55.5 seconds elapsed (6min 45s CPU)
using 8 threads. This required 11 lines of code, using only the 
Python standard library and NumPy. Similarly, computing allele
counts at each site required 4min 53s (25min 35s CPU) using a slightly
more complicated 16 line Python function.
Existing methods for working with FASTA data provided efficient 
access to individual alignments using the fai index format, which
works with uncompressed or bgzip compressed files.
For example, retrieving the alignment for the 1 millionth sample 
from the corresponding chunk file requires only a few milliseconds
using pyfaidx~\citep{shirley2015efficient} or 
pysam~\citep{pysam}. Retrieving the same alignment using Zarr 
required 365ms. 
Accessing calls for all samples at at a given 
site requires a similar time (here, accessing site 14,951
required 276ms). 
FASTA cannot support efficient retrieval
of data for a particular site in a set of alignments.
Storing data chunked in two dimensions means that we can avoid
the classic choice between sample-major or variant-major storage,
and access it efficiently in both dimensions.

>>>>

\textbf{Results:}
Zarr is a format for storing
multi-dimensional data that is widely used across the sciences,
and is ideally suited to massively parallel processing.
We present the VCF Zarr specification, an encoding of the
VCF data model using Zarr, along with fundamental software infrastructure
for efficient and reliable conversion at scale.
We show how this format is far more efficient than
standard VCF based approaches,
and competitive with specialised methods for
storing genotype data in terms of compression ratios
and single-threaded calculation performance.
We present case studies on subsets of three large 
human datasets (Genomics England: $n$=78,195;
Our Future Health: $n$=651,050; 
All of Us: $n$=245,394) along with whole genome
datasets for Norway Spruce ($n$=1,063) and 
SARS-CoV-2 ($n$=4,484,157).
We demonstrate the potential for VCF Zarr to enable a new generation
of high-performance and cost-effective applications
via illustrative examples using cloud computing and GPUs.

@@

In this article, we decouple the VCF data model from its row-oriented
file definition, and show how the data can be
compactly stored and efficiently analysed in a cloud-native, FAIR manner.
We do this by translating VCF data into Zarr format,
a method of storing large-scale multidimensional data as a regular
grid of compressed chunks.
Zarr's elegant simplicity and first-class support for
cloud object stores and parallel computing have led to
it gaining substantial traction
across the sciences, and it is now used in multiple petabyte-scale
datasets in cloud deployments (see Methods for details).
We present the VCF Zarr specification that formalises this
mapping, along with the \texttt{vcf2zarr} and \texttt{vcztools} 
utilities to reliably convert VCF to and from Zarr at scale.
We show that VCF Zarr is much more compact than
VCF and is competitive with state-of-the-art
file-based VCF compression tools.
Moreover, we show that Zarr's storage of data in an analysis-ready
format greatly facilitates computation,
with various benchmarks being substantially faster than
\texttt{bcftools} based pipelines, and again competitive
with state-of-the-art file-oriented methods. 
We demonstrate the scalability and flexibility of VCF Zarr
via case-studies on five boundary-pushing datasets.
We examine conversion and compression performance on three different
data modalities for large human datasets: whole genome sequence
data from Genomics England (5X smaller than gzipped VCF);
genotype data from Our Future Health (2.5X smaller than gzipped VCF);
and exome-like data from All of Us (1.1X larger than gzipped VCF). 
We then examine performance 
for species with large genomes using Norway Spruce data (3.2X smaller
than gzipped VCF),
and in a large collection of short whole genome alignments
using SARS-CoV-2 data 
(77X smaller than gzipped FASTA). 
Interoperability with the vast ecosystem of tools
based on VCF~\citep{danecek2021twelve,garrison2022spectrum}
is crucial. 
% TODO sentence too long
We first present the \texttt{vcztools} program,
which implements a subset of the functionality in \texttt{bcftools}
and allows data stored in Zarr at some URL to be 
treated as if it were a VCF or BCF file,
and demonstrate that it can be used as a practical replacement 
in processing pipelines.
We then present an illustrative example of how existing
methods can adopt VCF Zarr by modifying the popular 
SAIGE~\cite{zhou2018efficiently, zhou2020scalable} GWAS software
(written in C++), and show significant speedups over the existing 
VCF (7.9X) and BCF (3.4X) backends in a simple benchmark.
Finally, we illustrate the potential 
for a new generation of Zarr-based applications via some illustative
benchmarks. First, we demonstrate the very high levels of throughput 
performance available on public clouds when data in an object 
store is accessed in an asynchronous and highly parallel manner.
Finally, we demonstrate the ease and efficiency with which data
in Zarr format can be processed on a GPU, opening many exciting
avenues for future research.

@@

\subsection{Case study: SARS-CoV-2 alignment data}
The SARS-CoV-2 pandemic resulted in viral sequence data collection at
an unprecedented scale, with over 15 million whole genomes incorporated into
one resource as of 2024~\citep{hinrichs2024ongoing}. Data at such scale
overwhelmed classical phylogenetic methods, leading to the development
of a new class of tools~\citep{turakhia2021ultrafast,ye2022matoptimize,zhan2023towards}
capable of handling millions of samples.
The Viridian dataset~\citep{hunt2024addressing} is a consistently assembled
set of consensus sequences for 4,484,157 samples, addressing a range of systematic
errors. The \texttt{sc2ts} inference tool uses VCF-Zarr as its
input format, and Zhan et al have made the Viridian dataset,
aligned to the reference using MAFFT~\citep{katoh2013mafft},
available for download~\citep{zhan2023towards}. This dataset has quite
different properties to those examined in previous case studies, and
illustrates the flexibility of Zarr across a range of biological applications.

\begin{table}
\caption{Summary for VCF Zarr conversion of Viridian SARS-CoV-2
whole genome alignments and metadata 
(4,484,157 samples, 29,903 variants)
consisting of 37 fields and 401 MiB of storage, 
of which 279 is used by alignments ($\sim 77$X smaller 
than gzipped FASTA). The chunk size is 100 variants $\times$
10,000 samples.
Shown are the top 9 fields consuming at least 1\% of the 
overall storage (see Table~\ref{tab-genomics-england-data} for column details).
Field names (derived from source TSV file) are shortened for 
display purposes by replacing ``Genbank'' with ``G''
and ``Viridian'' with ``V''.
\label{tab-sc2-data}}
\begin{tabular}{llS[table-format=3.2]S[table-format=3.2]S[table-format=3.2]}
\toprule
{Field} & {Type} & {Storage} & {Compress} & {\% Total} \\
\midrule
call\_genotype & int8 & 279.38 MiB & 447.5 & 80.04\% \\
sample\_G\_tree\_name & str & 13.18 MiB & 2.5 & 3.78\% \\
sample\_G\_accession & str & 7.31 MiB & 4.6 & 2.09\% \\
sample\_Sample & str & 7.26 MiB & 4.6 & 2.08\% \\
sample\_Experiment & str & 6.9 MiB & 4.8 & 1.98\% \\
sample\_id & str & 6.83 MiB & 4.9 & 1.96\% \\
sample\_V\_N & int16 & 4.32 MiB & 1.9 & 1.24\% \\
sample\_V\_pangolin\_1.29 & str & 3.98 MiB & 8.4 & 1.14\% \\
sample\_V\_pangolin & str & 3.89 MiB & 8.6 & 1.11\% \\
\bottomrule
\end{tabular}
\end{table}

@@

Decompressed, the Viridian dataset consists of 48
FASTA files totalling 125 GiB of storage and a 1.4 GiB
TSV file for sample metadata.
When aligned to the reference and converted to Zarr, the
entire dataset is stored in a single 401 MiB Zip file.
When compressed with bgzip (so they can be indexed; see next 
paragraph) the alignments require 22 GiB of storage.
Table~\ref{tab-sc2-data} summarises the storage used by individual
arrays, and shows that the alignment data requires 279 MiB,
around 77X smaller than gzipped FASTA.
The original dataset is distributed as xz compressed FASTA files, which
achieves even higher compression levels at a total of 218 MiB.
However, these files must be decompressed in order to access the data
using standard FASTA processing techniques.

Computational accessibility for the Zarr is excellent. For example,
computing the total number of missing or ambiguous bases for 
all 4.4 million samples required 57.4 seconds elapsed (7min 08s CPU)
using 8 threads. This required 11 lines of code, using only the 
Python standard library and NumPy. Similarly, computing allele
counts at each site required 4min 19s (20min 50s CPU) using a slightly
more complicated 16 line Python function.
Existing methods for working with FASTA data provided efficient 
access to individual alignments using the fai index format, which
works with uncompressed or bgzip compressed files.
For example, retrieving the alignment for the 1 millionth sample 
from the corresponding chunk file requires only a few milliseconds
using pyfaidx~\citep{shirley2015efficient} or 
pysam~\citep{pysam}. 
Retrieving the same alignment using Zarr 
required 327ms. 
Accessing calls for all samples at at a given 
site requires a similar time (here, accessing site 14,951
required 482ms). 
FASTA cannot support efficient retrieval
of data for a particular site in a set of alignments.
Storing data chunked in two dimensions means that we can avoid
the classic choice between sample-major or variant-major storage,
and access it efficiently in both dimensions.
